---
title: "05_Shiny"
author: "Abby Lewis"
date: "2023-10-27"
output: html_document
---

# Shiny Apps 101

Shiny apps are interactive web applications built in R. Using Shiny apps can be a nice way to share your work with others, through dashboards, interactive visualizations, etc.

This tutorial will provide the basics of how to build a Shiny app. For more information, check out the [Shiny website](https://shiny.posit.co/).

This lesson also includes a little crash course on creating maps in R and creating interactive visualizations with plotly. For more mapping resources, [this](https://r-spatial.org/r/2018/10/25/ggplot2-sf.html) is my favorite reference.

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(tidyverse)
```

## Basic Shiny app

Here's a basic Shiny app to illustrate how the framework works
```{r}
# The UI is where you specify what you want the user to see
ui <- fluidPage(
  titlePanel("Basic Shiny App"), #Title
  
  selectInput(inputId = "stored_value", #Drop-down menu
              label = "This is a drop-down menu", 
              choices = c("Option 1", "Option 2")),
  
  textOutput(outputId = "text")
)

#The server is where you put behind-the-scenes code that contributes to the UI
server <- function(input, output, session) {
  
  output$text <- renderText({
    paste0("You selected: ", input$stored_value)
  })
  
}

#Running shinyApp creates the application using the ui and server
shinyApp(ui, server)
```

## Birds!

Lets get some data to work with. Here, I'm using tidytuesday bird counts

```{r}
feederwatch <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-01-10/PFW_2021_public.csv')
site_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-01-10/PFW_count_site_data_public_2021.csv')
species_translator <- readr::read_csv('https://feederwatch.org/wp-content/uploads/2022/08/PFW-species-translation-table.csv')

#Add date to feederwatch
feederwatch$Date <- as.Date(paste0(feederwatch$Year, "-", feederwatch$Month, "-", feederwatch$Day))
```

Before we turn this into an application, let's take a look at the data structure ourself

```{r}
feederwatch
site_data
species_translator
```

Quick visualization to see what we're working with

```{r}
#Add common names to feederwatch
feederwatch <- feederwatch %>%
  left_join(species_translator)

#How many observations of each species
observations <- feederwatch %>%
  group_by(species_code, american_english_name) %>%
  summarize(total = sum(how_many, na.rm = T)) 

observations %>%
  ungroup() %>%
  top_n(10, total) %>% #Lets only look at the top 10 most observed birds 
  ggplot(aes(x = total, y = american_english_name)) +
  geom_col() + 
  theme(axis.title.y = element_blank())  #supress y axis title
```

## Add data to shiny app

To add this figure to our application, we add `plotOutput()` to the UI, and `output$plot <- renderPlot({})` to the server.

Within renderPlot({}) we can paste the exact same figure code we used outside of the Shiny application

```{r}
ui <- fluidPage(
  #Add title
  titlePanel("Plot of total observations"),
  
  #Add plot
  plotOutput("plot")
  )

server = function(input, output, session) {
  #Generate plot on the server side
  output$plot <- renderPlot({
    
    #Copy and paste figure code from above (no edits needed!)
    
    })
}

shinyApp(ui, server)
```

### Add interactivity

What if we want users to be able to control how many birds show up?

We do this by adding `sliderInput()` to the UI, and using `input$n` to control the number of birds in the plot in the server code

```{r}
ui <- fluidPage(
  #Add title
  titlePanel("Plot of total observations"),
  
  #Add slider
  sliderInput(inputId = "n", 
              label = "Number to plot", 
              min = 3, max = 30, value = 10),
  
  #Add plot
  plotOutput("plot")
  )

server = function(input, output, session) {
  
  #Generate plot on the server side
  output$plot <- renderPlot({
    
    feederwatch %>%
      group_by(species_code, american_english_name) %>%
      summarize(total = sum(how_many, na.rm = T)) %>%
      ungroup() %>%
      top_n(input$n, total) %>% #Add input$n here
      ggplot(aes(x = total, y = american_english_name)) +
      geom_col() + 
      theme(axis.title.y = element_blank()) 
    
    })
}

shinyApp(ui, server)
```

## Plotly

Another way to add interactivity to plots is with plotly. For example, using plotly, we can hover to see the exact number of observations for each bird

```{r}
#Load plotly
library(plotly)

#Create plot
plot <- feederwatch %>%
  group_by(species_code, american_english_name) %>%
  summarize(total = sum(how_many, na.rm = T)) %>%
  ungroup() %>%
  top_n(10, total) %>% #Lets only look at the top 10 most observed birds 
  ggplot(aes(x = total, y = american_english_name)) +
  geom_col() + 
  theme(axis.title.y = element_blank())  #supress y axis title

#Convert to plotly
ggplotly(plot)
```

To incorporate this into a shiny app, we use `plotlyOutput()` in the UI, and `renderPlotly()` in the server.

```{r}
ui <- fluidPage(
  titlePanel("Plot of total observations"),
  
  sliderInput(inputId = "n", 
              label = "Number to plot", 
              min = 3, max = 30, value = 10),
  
  #plotlyOutput instead of plotOutput
  plotlyOutput("plot")
  )

server = function(input, output, session) {
  #renderPlotly instead of renderPlot
  output$plot <- renderPlotly({

    plot <- feederwatch %>%
      group_by(species_code, american_english_name) %>%
      summarize(total = sum(how_many, na.rm = T)) %>%
      ungroup() %>%
      top_n(input$n, total) %>% #Add input$n here
      ggplot(aes(x = total, y = american_english_name)) +
      geom_col() + 
      theme(axis.title.y = element_blank())
    
    ggplotly(plot)
})
}

shinyApp(ui, server)
```

## Mapping

Dark eyed juncos are the most observed bird in the dataset. Let's look at how frequently they are observed at different sites.

To do so wil require a good bit of work. Here are the steps:
1. Calculate whether dark eyed juncos have been observed at each site (practice data wrangling)
2. Create a basic plot (practice with ggplot syntax)
3. Turn our basic plot into a pretty map (practice making maps in R)
4. Add the map to a Shiny App, adding interactivity (practice with Shiny)

### Step 1: Calculate presence

The dataset only includes indications of where species _were_ observed, and does not explicitly specify where they were not observed. Let's do some data processing to calculate presence across the different sites.

```{r}
#First, explicitly fill in NAs in the dataset. We want all combinations of sites and species
loc_id <- rep(unique(feederwatch$loc_id), 
             each = length(unique(feederwatch$species_code)))
species_code <- rep(unique(feederwatch$species_code), 
               times = length(unique(feederwatch$loc_id)))
NA_filled_sites <- data.frame(loc_id, species_code) 

#Then, re-add the bird observation data
NA_filled_data <- NA_filled_sites %>% 
  left_join(feederwatch %>% 
              select(loc_id, species_code, Date, how_many)) 

#And calculate whether the bird has been observed at each site 
#(this will take a few seconds)
presence <- NA_filled_data %>%
  group_by(loc_id, species_code) %>%
  summarize(n = sum(how_many, na.rm = T),
            present = n > 0 & !is.na(n)) 

#Fill in lat/long for plot
#Identify unique lat/long
latlongs <- feederwatch %>% 
  select(loc_id, latitude, longitude) %>%
  unique()
#Add to presence
presence_latlong <- presence %>%
  left_join(latlongs)

#Add common names
presence_latlong <- presence_latlong %>%
  left_join(species_translator)
```

### Step 2: Basic plot

Now we're ready to plot!
```{r}
#Plot
presence_latlong %>%
  filter(species_code == "daejun") %>% #Select the species we want
  ggplot(aes(x = longitude, y = latitude)) + #Using lat/long as x/y
  #Plotting points in two steps so presence is on top of absence
  geom_point(data = . %>% filter(!present), #Absence
             aes(fill = present), 
             shape = 21, #outlined circle
             alpha = 0.2 #make absences transparent
             ) + 
  geom_point(data = . %>% filter(present), #Presence
             aes(fill = present), 
             shape = 21) +
  #Set colors
  scale_fill_manual(values = c("white", "blue"))
```

This is nice, but a pretty bad looking map. Let's use the sf package to make it better.

### Step 3: Pretty map!

Load packages and base maps.

```{r}
#Load packages
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

#Load base maps
states <- ne_states(returnclass = "sf", country = "United States of America")
world <- ne_countries(scale = "medium", returnclass = "sf")

#Filter data for this species
for_map <- presence_latlong %>%
      filter(species_code == "daejun")
```

Then, create plot.

Here I am layering state boundaries on top of country boundaries to also include Canada. The rest of the code should look quite similar to the basic map we made above.

```{r}
#Create plot
ggplot() +
  geom_sf(data = world) + #Add country boundaries
  geom_sf(data = states) + #Add state boundaries
  coord_sf(expand = FALSE, ylim = c(23, 55), xlim = c(-130,-60))+
  geom_point(aes(fill = present, 
                 y = latitude, 
                 x = longitude), 
             shape = 21, #outlined circle
             alpha = 0.3, #make absences transparent
             data = for_map %>% filter(!present)) + 
  geom_point(aes(fill = present, 
                 y = latitude, 
                 x = longitude), 
             shape = 21,
             data = for_map %>% filter(present)) +
  scale_fill_manual(values = c("white", "blue"), name = "Observed?") +
  theme_bw()+
  theme(plot.margin = margin(0, 0, 0, 0, "cm"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank())
```

### Step 4: Add to Shiny App

You know the drill- just paste into our existing shiny app framework! 

Only the parts that are dynamic need to be in the server—for the rest (e.g., base maps), it is sufficient we that they are in our global environment because we created these above.

```{r}
ui <- fluidPage(
  #Add drop down menu for species
  selectInput("species_name", "Select species", 
              choices = unique(feederwatch$american_english_name)),
  
  #Add plot
  plotOutput("plot")
)

server = function(input, output, session) {
  output$plot <- renderPlot({
    
    #Filter data for the input species
    for_map <- presence_latlong %>%
          filter(american_english_name == input$species_name)
    
    #Create plot
    ggplot() +
      geom_sf(data = world) + 
      geom_sf(data = states) + 
      coord_sf(expand = FALSE, ylim = c(23, 55), xlim = c(-130,-60))+
      geom_point(aes(fill = present, 
                     y = latitude, 
                     x = longitude), 
                 shape = 21, 
                 alpha = 0.3, 
                 data = for_map %>% filter(!present)) + 
      geom_point(aes(fill = present, 
                     y = latitude, 
                     x = longitude), 
                 shape = 21,
                 data = for_map %>% filter(present)) +
      scale_fill_manual(values = c("white", "blue"), name = "Observed?") +
      theme_bw()+
      ggtitle(input$species_name)+ #Add title with input species name
      theme(plot.margin = margin(0, 0, 0, 0, "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
    
    })
}

shinyApp(ui, server)
```

## Your turn!

BEGINNER: Let's practice modifying a basic Shiny app, setting aside the data for now.

Customize this UI to include different text (i.e., replacing "Hello, world!") and a different control widget (i.e., replacing `selectInput()`). The control widget won't do anything yet, but that's okay! This is just practice.

For more information on what control widgets are available in Shiny, see [here](https://shiny.posit.co/r/getstarted/shiny-basics/lesson3/).

```{r}
# The UI is where you specify what you want the user to see
ui <- fluidPage(
  "Hello, world!",
  
  selectInput("stored_value", "This is a drop-down menu", 
              choices = c("Option 1", "Option 2")),
)

#The server is where you put behind-the-scenes code that contributes to the UI
#No server specifications needed for this very basic application
server <- function(input, output, session) {
}

#Running shinyApp creates the application using the ui and server
shinyApp(ui, server)
```

INTERMEDIATE: I'm curious how the relative prevalence of different species changes across months of the year. 

Modify this code to allow users to select what species they want on the plot. 

I've done the data wrangling for you in the code chunk below—you just need to fill in the shiny app details.

```{r}
#Data wrangling
#Calculate the percent of all observations each month that are from a given species (done)
percent_per_month <- feederwatch %>%
  group_by(Year, Month) %>%
  mutate(tot = sum(how_many, na.rm = T)) %>%
  group_by(Year, Month, species_code) %>%
  summarize(n = sum(how_many, na.rm = T),
            pct = n/unique(tot) * 100) %>%
  mutate(Date = as.Date(paste0(Year, "-", Month, "-01"))) %>%
  left_join(species_translator)

#Get the 20 most common species to include in our checklist (done)
most_common <- feederwatch %>%
  group_by(species_code, american_english_name) %>%
  summarize(total = sum(how_many, na.rm = T)) %>%
  ungroup() %>%
  top_n(20, total)
species_options <- unique(most_common$american_english_name)



#SKELETON STRUCTURE FOR SHINY APP- 
#It is your job to fill this in

ui <- fluidPage(
  #Add drop down menu for species using checkboxGroupInput()
  #See the help page for more instructions
  
  #Add plot
)

server = function(input, output, session) {
  output$plot <- renderPlot({
    
    percent_per_month %>%
      filter(american_english_name %in% XXX ) #(fill in XXX in this line to respond to the UI)
      ggplot(aes(x = Date, y = pct, color = american_english_name)) +
      geom_line()+
      labs(color = "Species") +
      ylab("Percent of all observed birds")+
      xlab("Month")
    
    })

}

shinyApp(ui, server)
```

ADVANCED:

What are _you_ interested in? Create a Shiny app to explore the data using the space below.

```{r}

```

