---
title: "Function practice"
author: "Abby Lewis"
date: "2023-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Function practice
Primarily from [R-bloggers](https://www.r-bloggers.com/2016/02/functions-exercises/)

## Exercise 1
Create a function `sum2()` that will return the sum of 2 integers.
* E.g., `sum2(1,3)` should return 4

```{r}

```

## Exercise 2
Create a function that will return TRUE if a given integer is inside a vector.
* E.g., `is_inside(1, c(1, 3, 6, 7))` should return TRUE

```{r}

```

## Exercise 3
Create the function `unique_elements()`, which given a vector will return a vector with the duplicated elements from the input vector removed.
* E.g., `unique_elements(c(1, 3, 6, 7, 1, 3, 6))` should return c(1, 3, 6, 7)

```{r}

```

## Exercise 4
Create a function `summarize_stats()` that given a vector will print by screen the mean and the standard deviation, it will optionally also print the median.
* E.g., `summarize_stats(c(1, 3, 6, 7, 1, 3, 6))` should print a mean of 3.857143 and an SD of 2.478479
* `summarize_stats(c(1, 3, 6, 7, 1, 3, 6), median = T)` should also print a median of 3

```{r}

```

## Exercise 5
Create a function `guess_number()` that takes an input from the user using `readline()` and returns yes or no depending upon whether the user correctly guesses the secrete number 7.

BONUS: modify this function to use a "while" loop and continue taking inputs until the user correctly guesses the number
```{r}
guess_number <- function(){
  message("Guess a number: ")
  guess <- readline()
  
  # Additional code here
}

guess_number()
```

