---
title: "Text Processing and Regular Expressions"
output: html_document
date: "2023-11-08"
author: Megan Moran
---

This document is modified from Chapter 1.7 of [Mastering Software Development in R](https://bookdown.org/rdpeng/RProgDA/text-processing-and-regular-expressions.html) by by Roger D. Peng, Sean Kross and Brooke Anderson .

Regular Expressions

A regular expression is a string that defines a pattern that could be contained within another string. A regular expression can be used for searching for a string, searching within a string, or replacing one part of a string with another string.

We'll be using the grepl() function, which takes two arguments: a regular expression and a string to be searched. If the string contains the specified regular expression then grepl() will return TRUE, otherwise it will return FALSE

```{r}
grepl("land", "Maryland")
```

Metacharacters can be used to look for specific parts of strings. A period "." represents any character other than a new line.

```{r}
grepl("a.b", c("aaa", "aab", "abb", "acadb"))
```

The + metacharacter indicates that one or more of the preceding expression should be present and \* indicates that zero or more of the preceding expression is present.

```{r}
# Does "Maryland" contain zero or more of "a" ?
grepl("a*", "Maryland")

# Does "Maryland" contain one or more of "x" ?
grepl("x+", "Maryland")

```

You can also specify exact numbers of expressions using curly brackets {}.

```{r}
# Does "Mississippi" contain exactly 2 adjacent "s" ?
grepl("s{2}", "Mississippi")

# Does "Mississippi" contain between 1 and 3 adjacent "s" ?
grepl("s{1,3}", "Mississippi")


```

Parentheses () can be used to create a capturing group. A capturing group allows you to use quantifiers on other regular expressions.

```{r}
# Does "Mississippi" contain the pattern of an "i" followed by 
# 2 of any character, with that pattern repeated three times adjacently?
grepl("(i.{2}){3}", "Mississippi")
```

You can specify sets of characters with regular expressions: words ("\\w") specify any letter digits ("\\d") specify digits 1 through 9 whitespace characters ("\\s") specifies line breaks, tabs, or spaces. Each of these character sets have their own compliments: not words ("\\W"), not digits ("\\D"), and not whitespace characters ("\\S"). Each specifies all of the characters not included in their corresponding character sets.

```{r}
grepl("\\w", "abcdefghijklmnopqrstuvwxyz0123456789")
grepl("\\d", "0123456789")
```

You can also specify specific character sets using straight brackets []. For example a character set of just the vowels would look like: "[aeiou]".

You can find the complement to a specific character by putting a carrot \^ after the first bracket. For example "[\^aeiou]" matches all characters except the lowercase vowels.

You can specify ranges of characters using a hyphen - inside of the brackets. For example "[a-m]" matches all of the lowercase characters between a and m, while "[5-8]" matches any digit between 5 and 8 inclusive.

```{r}
grepl("[^aeiou]", "rhythms")
grepl("[a-m]", "xyz")
```

Putting two backslashes before a punctuation mark that is also a metacharacter indicates that you are looking for the symbol and not the metacharacter meaning. For example "\\." indicates you are trying to match a period in a string.

```{r}
grepl("\\+", "tragedy + time = humor")
```

Metacharacters for matching the beginning and the end of a string are "\^" and "\$" respectively.

```{r}
grepl("^a", c("bab", "aab"))
grepl("b$", c("bab", "aab"))
```

The OR ("\|") metacharacter matches either the regular expression on the left or on the right side of this character.

```{r}
grepl("a|b", c("abc", "bcd", "cde"))
```

Let's create a regular expression that matches all state names (using the R dataset state.names) that both begin and end with a vowel:

1.  We match the beginning of a string.
2.  We create a character set of just capitalized vowels.
3.  We specify one instance of that set.
4.  Then any number of characters until:
5.  A character set of just lowercase vowels.
6.  We specify one instance of that set.
7.  We match the end of a string.

```{r}
start_end_vowel <- "^[AEIOU]{1}.+[aeiou]{1}$"
vowel_state_lgl <- grepl(start_end_vowel, state.name)
head(vowel_state_lgl)

state.name[vowel_state_lgl]
```
