---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
Install necessary packages

Load needed package
```{r}
library(lidR)
library(raster)
library(tinytex)
```
The function readLAS() is used to read and import LiDAR data files in the LAS/LAZ (LiDAR Data Exchange Format) format. LAS/LAZ files store point cloud data generated by LiDAR sensors, including information about the x, y, and z coordinates of individual points, as well as additional attributes such as intensity and classification.
```{r}
#read LAZ file
las <- readLAS('neon_laz.laz')
```
print summary of the file
```{r}
print(las)
```
Check the validation of the data
```{r}
las_check(las)
```
Plot 3D lidar plot
```{r}
plot(las)
```
Make Canopy height model
```{r}
chm <- rasterize_canopy(las, res = 1, algorithm = p2r())
col <- height.colors(25)
color_palette <- colorRampPalette(col)
#plot(chm, col = col)
plot(chm, col = color_palette(25))
```
classify the points
```{r}
class <- classify_ground(las, algorithm = pmf(ws = 5, th = 3))

```
```{r}
plot(class, color = "Classification", size = 3, bg = "white")
```

